---
title: "Clustering trajectories"
author: "Hatfield, Heggeseth, Schuler"
date: "October 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width = 10
)
library(biclust) # loads MASS, which has select that will mask in dplyr
library(ggplot2)
library(gridExtra) # masks combine
library(dplyr)
library(tidyr) 
library(readr)
library(directlabels)
library(knitr)
library(igraph) # masks distances
library(distances)
library(cluster)
library(mclust)

load("synthetic_ESSCLC_data.RData")
theme_set(theme_minimal())
max.days <- 365

## Function to make the plots
## takes sorted trajectory data
plot.sorted.trajectories <- function(traj.dat){
  ## Only plot 1 thru 100
  plot.scaled.traj <- ggplot(filter(traj.dat,ID%in%1:100),
                             aes(y=ID,yend=ID,x=FromDateRel,xend=ThruDateRel,group=ID)) +
  geom_segment(aes(col=setting),size=3) +
  xlab("Proportion") + ylab("") +
  scale_color_brewer(type="qual",palette=3,guide="none")+
  theme(legend.position="bottom",legend.title=element_blank(),panel.grid=element_blank(),axis.text.x=element_text(angle=90)) + coord_flip()
plot.traj <- ggplot(filter(traj.dat,ID%in%1:100),
                    aes(y=ID,yend=ID,x=FromDate,xend=ThruDate,group=ID)) +
  geom_segment(aes(col=setting),size=3) +
  scale_x_continuous("Days",limits=c(0,max.days)) + ylab("") +
  scale_color_brewer(type="qual",palette=3,guide="none")+
  theme(legend.position="bottom",legend.title=element_blank(),panel.grid=element_blank(),axis.text.x=element_text(angle=90)) + coord_flip()
grid.arrange(plot.scaled.traj,plot.traj)
}

## Function to make class-specific group time trajectories
## takes trajectory data object with cluster variable
plot.cluster.means <- function(traj.dat){
  cluster.labels <- select(traj.dat,ID,cluster) %>% group_by(ID) %>% slice(1) %>% 
    group_by(cluster) %>% summarize(count=n()) %>% ungroup() %>% 
    mutate(total=sum(count),prop=count/total*100,
           label=paste("Cluster ",cluster," (",round(prop,0),"%)",sep=""))
  resolution <- .1
  class.segment.dat <- traj.dat %>%
    select(cluster,setting,FromDateRel,ThruDateRel) %>%
    ## Segments are in increments of .05 of relative length:
    mutate(segment.min=round(FromDateRel / resolution,0),
           segment.max=round(ThruDateRel / resolution,0))
  class.summaries <- NULL
  for(i in 1:(1/resolution)) {
    these <- filter(class.segment.dat,i>=segment.min,i<=segment.max)
    temp <- left_join(these %>% group_by(cluster) %>% summarize(denom=n()),
                      these %>% group_by(cluster,setting) %>%
                        summarize(num=n()),by="cluster") %>%
      spread(key=setting,value=num) %>% 
      mutate_each(funs(ifelse(is.na(.),0,.)),-cluster,-denom) %>%
      gather(key=setting,value=num,-cluster,-denom) %>%
      mutate(prop=num/denom*100,segment=i) %>% arrange(cluster,setting)

    class.summaries <- bind_rows(class.summaries,temp) %>%
      mutate(cluster.label=factor(cluster,levels=cluster.labels$cluster,
                                  labels=cluster.labels$label))
    rm(these,temp)
  }
  ## Add zeros to empty segments (for small classes, especially)

  ggplot(class.summaries,aes(x=segment,y=prop,group=setting)) + 
    geom_smooth(aes(col=setting),se=F,method="loess") + facet_wrap(~cluster.label) +
    scale_color_brewer(palette="Dark2",guide="none") +
    theme(strip.background=element_blank(),panel.grid=element_blank(),
          axis.text=element_blank()) +
    scale_y_continuous("",limits=c(0,100)) + 
    scale_x_continuous("",expand=c(0,1.1))  +
    geom_dl(aes(label=setting,col=setting),method=list("top.bumpup",cex=.8,hjust=.5))
}
```

## Sorted on true class and then ID (i.e., arbitrary within class)
```{r plot_trajectories}
## Subset to a small number of people
these.episodes <- filter(episodes,ID%in%1:200) %>% ungroup() %>%
  left_join(summary.stats %>% select(ID,class),by="ID")
first.ep <- these.episodes %>% group_by(ID,setting) %>% slice(1) %>%
  select(ID,setting,FromDateRel) %>%
  spread(key=setting,value=FromDateRel) %>%
  rename(first.Hospice=Hospice,first.Hospital=Hospital,
         first.NoCoveredCare=NoCoveredCare,
         first.SNF=SNF) %>% ungroup()
these.sumstats <- filter(summary.stats,ID%in%1:200) %>% ungroup() %>%
  left_join(first.ep,by="ID") %>%
  mutate_each(funs(ifelse(is.na(.),0,.)),first.Hospice:first.SNF)

## Plot the new trajectories first 100 records (used subsequently)
true.traj.dat <- these.episodes %>%
  mutate(ID=factor(ID,levels=order(summary.stats$class,summary.stats$ID)),
         cluster=class)

plot.sorted.trajectories(true.traj.dat)
plot.cluster.means(true.traj.dat)
```

## Divisive clustering on Gower's distance measure from original summary stats (standardized)
```{r orig_summaries}
orig.dmat <- daisy(these.sumstats %>%
                     select(perc.Hospice:SNF.Count),
                   metric="gower",stand=T)
orig.clusts <- diana(orig.dmat)

## Look at the trajectories in order of agglomerative clustering:
## Join back to the data and plot
orig.traj.dat <- these.episodes %>%
  left_join(data.frame(ID=these.sumstats$ID,
                          cluster=cutree(as.hclust(orig.clusts),h=.35)),by="ID") %>%
  mutate(ID=factor(ID,levels=orig.clusts$order,ordered=T))

plot.sorted.trajectories(orig.traj.dat)

## Cut them into classes and plot the average trajectories by class:
pltree(orig.clusts,main="",xlab="",sub="");abline(h=.35,lty=2)
plot.cluster.means(orig.traj.dat)
```

## Divisive clustering on Gower's distance measure from original summary stats plus survival time (standardized)
```{r orig_plus_surv}
surv.dmat <- daisy(these.sumstats %>%
                     select(perc.Hospice:SNF.Count,Survival.Days),
                   metric="gower",stand=T)
surv.clusts <- diana(surv.dmat)

## Look at the trajectories in order of agglomerative clustering:
## Join back to the data and plot
surv.traj.dat <- these.episodes %>%
  left_join(data.frame(ID=these.sumstats$ID,
                          cluster=cutree(as.hclust(surv.clusts),h=.35)),by="ID") %>%
  mutate(ID=factor(ID,levels=surv.clusts$order,ordered=T))

## Cut them into classes and plot the average trajectories by class:
pltree(surv.clusts,main="",xlab="",sub="");abline(h=.35,lty=2)
plot.cluster.means(surv.traj.dat)
```

## Divisive clustering on Gower's distance measure from original summary stats plus survival time and relative timing of first episode in each setting (standardized)
```{r orig_plus_timing}
timing.dmat <- daisy(these.sumstats %>%
                    select(perc.Hospice:SNF.Count,Survival.Days,
                           first.Hospice:first.SNF),
                    metric="gower",stand=T)
timing.clusts <- diana(timing.dmat)

## Look at the trajectories in order of agglomerative clustering:
## Join back to the data and plot
timing.traj.dat <- these.episodes %>%
    left_join(data.frame(ID=these.sumstats$ID,
                          cluster=cutree(as.hclust(timing.clusts),h=.35)),by="ID") %>%
  mutate(ID=factor(ID,levels=timing.clusts$order,ordered=T))

plot.sorted.trajectories(timing.traj.dat)

## Cut them into classes and plot the average trajectories by class:
pltree(timing.clusts,main="",sub="",xlab="");abline(h=.35,lty=2)
plot.cluster.means(timing.traj.dat)
```

## Divisive clustering on Gower's distance measure from original summary stats plus survival time and relative timing of first episode in each setting and site of death (standardized)
```{r orig_plus_death}
death.dmat <- daisy(these.sumstats %>%
                    select(perc.Hospice:SNF.Count,Survival.Days,
                           first.Hospice:first.SNF,site.death),
                    metric="gower",stand=T)
death.clusts <- diana(death.dmat)

## Look at the trajectories in order of agglomerative clustering:
## Join back to the data and plot
death.traj.dat <- these.episodes %>%
  left_join(data.frame(ID=these.sumstats$ID,
                       cluster=cutree(as.hclust(death.clusts),h=.35)),by="ID") %>%
  mutate(ID=factor(ID,levels=death.clusts$order,ordered=T))

plot.sorted.trajectories(death.traj.dat)

## Cut them into classes and plot the average trajectories by class:
pltree(death.clusts,main="",sub="",xlab="");abline(h=.35,lty=2)
plot.cluster.means(death.traj.dat)
```

